Data Specialization for a Staged Lambda Calculus

### 2014-11-08 paper/splitting.tex
Making text consistent with new terminology.

% Intuitively, we require that the terms obtained by splitting a term $e$, when
% evaluated, agree with $e$ when evaluated according to our dynamic semantics. In
% the case of stage~\bbone\ terms, our dynamic semantics produce partial values
% containing information at \emph{both} stages, which goes against the idea that
% splitting separates the subcomputations belonging to each stage.

Recall that \lang\ partial values and residual tables together may
contain content at \emph{both} stages. For example the term:
\begin{lstlisting}
1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
\end{lstlisting}
evaluates via $\redonesym$ to:
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}
Intuitively, splitting this term should generate an~$e_1$ that reduces to the stage~\bbone\ content of the partial value \texttt{3},
and an~$e_2$ corresponding to the stage~\bbtwo\ computation: \\
$\letin \yhat {\mathtt{7+5}} \yhat$.

To precisely specify the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
\emph{masking} functions, written $\masko{{-};{-}}$ and $\maskt{{-};{-}}$, 
which take a residual table and partial value as input and emit a term in \langmono\
that corresponds to only the content belonging to a single stage. \ref{fig:valMask} defines these
masking functions.

\emph{Stage~\bbone\ masking} of the partial value~$v$ (written $\masko{v}$) produces a term in \langmono\ containing exactly
the stage~\bbone\ content of~$v$. Since \next\ terms only contain stage~\bbtwo\
content, $\masko{v}$ replaces all $\next$ subterms of~$v$ with \texttt{()}.  Since lambdas may represent multi-stage computations, stage~\bbone\ masking splits their bodies as general stage~\bbone\ terms (as described in \ref{sec:split-one}), and returns
the stage~\bbone\ component. $\masko{v}$ always produces a value in \langmono\ that has
the same structure as~$v$.  For example, the stage~\bbone\ mask of the partial value given above is \texttt{(3,())}.

\emph{Stage~\bbtwo\ masking} of the partial value $v$ and residual context~$\xi$ (written $\maskt{\xi;v}$) produces a term in \langmono\ containing
exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike stage~\bbone\ masking, stage~\bbtwo\ masking 
requires access to the residual table $\xi$ because all stage~\bbtwo\
content in~$v$ has been lifted out into~$\xi$. Stage~\bbtwo\ masking discards
stage~\bbone\ content in~$v$ by replacing base constants with~\texttt{()},
replacing $\next~\hat{y}$ with~$\hat{y}$, and replacing all $n$-ary term
constructors with an $n$-tuple of masked subterms. Like stage~\bbone\ masking, stage~\bbtwo\
masking splits the bodies of lambdas as general stage~\bbone\ terms, and
returns the stage~\bbtwo\ component. After discarding all stage~\bbone\ content from~$v$, stage~\bbtwo\ masking reifies $\xi$ around the result,
producing a term in \langmono. For example, the stage~\bbtwo\ mask of the partial value and residual table given above is:
\begin{lstlisting}
let yhat=7+5 in ((),yhat)
\end{lstlisting}

>>>>

Recall that partial values and residual tables produced by \bbone-evaluation may
contain content at \emph{both} stages. For example the term:
\begin{lstlisting}
1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
\end{lstlisting}
evaluates via $\redonesym$ to:
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}
To match the results of \bbone-evaluation,
splitting the term should generate an~$e_1$ that reduces to the stage~\bbone\ content of the partial value, \texttt{3},
and an~$e_2$ corresponding to the stage~\bbtwo\ computation: \\
$\letin \yhat {\mathtt{7+5}} \yhat$.

to precisely specific the stage~\bbone\ and stage~\bbtwo\ components of \bbone-evaluation results, we define 
\emph{masking} functions, written $\masko{{-};{-}}$ and $\maskt{{-};{-}}$, 
which take a residual table and partial value as input and emit a term in \langmono\
that corresponds to only the content belonging to a single stage.
\ref{fig:valMask} defines these masking functions.

\emph{Stage~\bbone\ masking} of the partial value~$v$ (written $\masko{v}$) produces a term in \langmono\ containing exactly
the stage~\bbone\ content of~$v$. Since \next\ terms only contain stage~\bbtwo\
content, $\masko{v}$ replaces all $\next$ subterms of~$v$ with \texttt{()}.  Since lambdas may represent multi-stage computations, stage~\bbone\ masking splits the body of lambdas as a general stage~\bbone\ term (as described in \ref{sec:split-one}), and returns
the stage~\bbone\ component. $\masko{v}$ always produces a value in \langmono\ that has
the same structure as~$v$.  For example, the stage~\bbone\ mask of the partial value given above is \texttt{(3,())}.

\emph{Stage~\bbtwo\ masking} of the partial value $v$ and residual context~$\xi$ (written $\maskt{\xi;v}$) produces a term in \langmono\ containing
exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike stage~\bbone\ masking, stage~\bbtwo\ masking 
requires access to the residual table $\xi$ because after \bbone-evaluation, all stage~\bbtwo\
has been lifted out of~$v$. Stage~\bbtwo\ masking discards
stage~\bbone\ content in~$v$ by replacing base constants with~\texttt{()},
replacing $\next~\hat{y}$ with~$\hat{y}$, and replacing all $n$-ary term
constructors with an $n$-tuple of masked subterms. Like stage~\bbone\ masking, stage~\bbtwo\
masking splits the bodies of lambdas as general stage~\bbone\ terms, and
returns the stage~\bbtwo\ component. After discarding all stage~\bbone\ content from~$v$, stage~\bbtwo\ masking reifies $\xi$ around the result,
producing a term in \langmono. This term reduces to a value that with the same structure as~$v$.
For example, the stage~\bbtwo\ mask of the partial value and residual table given above is:
\begin{lstlisting}
let yhat=7+5 in ((),yhat)
\end{lstlisting}
